---
title: "5problemset"
author: "Paul Garces"
date: "2023-03-07"
output: html_document
---

```{r read packages}
library(ggplot2)
library(dplyr)
library(tidyverse)
```

# Question 1
### 1.
```{r}
openfile <- read_delim("gapminder.csv")
openfile
```

### 2.
```{r}
ncol(openfile)
nrow(openfile)
```

### 3.
```{r}
head(openfile, 3)
```

The data looks fine, there are some NA's however, but they do not really affect the data.
# Question 2
### 1.
```{r question 2}
n_distinct(openfile$name, na.rm = TRUE)
n_distinct(openfile$iso3, na.rm = TRUE)
n_distinct(openfile$iso2, na.rm = TRUE)
```

### 2.
#### a
```{r}
openfile %>% filter(is.na(iso2)) %>% group_by(name) %>% distinct(iso2)
```

#### b
```{r}
openfile %>% filter(is.na(name)) %>% group_by(name) %>% distinct(iso3)
```

### 3.
```{r}
min(openfile$time, na.rm = TRUE)
max(openfile$time, na.rm = TRUE)
```

# Question 3
### 1.
```{r}
openfile %>% group_by(time) %>% summarize(na_co2 = sum(is.na(co2)),
                                          na_co2_PC = sum(is.na(co2_PC))) %>% arrange(desc(na_co2), na_co2_PC)
```
The years with the most missing values were 2017, 2018, and 2019.

### 2.
```{r}
usa <- openfile %>% filter(name == 'United States of America') 
china <- openfile %>% filter(name == 'China')
india <- openfile %>% filter(name == 'India')
argentina <- openfile %>% filter(name == "Argentina")
ecuador <- openfile %>% filter(name == "Ecuador")

  
ggplot(data = openfile) + 
  geom_line(data = usa, aes(x = time, y = co2, color = "USA")) +
  geom_line(data = china, aes(x = time, y = co2, color = "China")) +
  geom_line(data = india, aes(x = time, y = co2, color = "India")) +
  geom_line(data = argentina, aes(x = time, y = co2, color = "Argentina")) +
  geom_line(data = ecuador, aes(x = time, y = co2, color = "Ecuador"))
```

It can be noted that China has had the biggest increase in CO2 emission given the certain time frame, with the 

United States steadily keeping up, even though it is not as much as China. India has also seen an increase in 

CO2 emissions after having a steady increase as well. Ecuador and Argentina have had a relatively slow increase

in CO2 emission compared to the other countries in the graph.

### 3.
```{r}
ggplot(data = openfile) + 
  geom_line(data = usa, aes(x = time, y = co2_PC, color = "USA")) +
  geom_line(data = china, aes(x = time, y = co2_PC, color = "China")) +
  geom_line(data = india, aes(x = time, y = co2_PC, color = "India")) +
  geom_line(data = argentina, aes(x = time, y = co2_PC, color = "Argentina")) +
  geom_line(data = ecuador, aes(x = time, y = co2_PC, color = "Ecuador")) 
```

When looking at CO2 emissions per capita, it can be noted that the United States is the biggest producer between all countries.

However, while Argentina is in second for most CO2 emissions per capita, it must also be observed that China has the largest increase.

Ecuador is in a close third place and India, while increasing produces the least amount of emissions per capita.

### 4.
```{r}
openfile %>% filter(!is.na(region), time == "2016") %>% group_by(region) %>% summarize(avgco2PC = mean(co2_PC, na.rm = TRUE))
```

Based on the graph, it can be noted that Asia and Europe are both the top producers of CO2 per capita between all continents. 

Africa being the lowest between all that are listed and the Americas being third and Oceania being 4.

### 5.
``` {r}
openfile %>% filter(!is.na(region), time == "1960" | time == "2016") %>% group_by(region, time) %>%
  summarize(avgco2PC = mean(co2_PC, na.rm  = TRUE)) %>% ggplot(aes(region, avgco2PC, fill = factor(time))) +
  geom_bar(stat = "identity", position = position_dodge())
```

### 6.
```{r}
openfile %>%
  filter(!is.na(co2_PC), time == "2016", !is.na(name), !is.na(region)) %>%
  group_by(region, name) %>%
  arrange(region, co2_PC) %>%
  group_by(region) %>%
  filter(rank(co2_PC) <= 3 | rank(desc(co2_PC)) <= 3) %>% 
  select(region, name, co2_PC)
```

# Question 4
### 1.
```{r}
gdple1960 <- openfile %>% 
  filter(!is.na(GDP_PC), !is.na(lifeExpectancy), time =="1960", !is.na(region)) %>% 
  ggplot(aes(x = GDP_PC, y = lifeExpectancy)) +
  geom_point(aes(size = totalPopulation, color = region))
gdple1960
```

There does appear to be a relationship between GDP per capita in a country and the life expectancy, which means that the higher the GDP per capita, the higher the life expectancy.

It can be noted that Africa and Asia had relatively low GDP per capita and life expectancy, while the opposite is the case for Europe and Asia.

It also appears that medium to small size countries have a higher GDP per capita and life expectancy when compared to larger ones.

### 2.
```{r}
gdple2019 <- openfile %>% 
  filter(!is.na(GDP_PC), !is.na(lifeExpectancy), time =="2019", !is.na(region)) %>% 
  ggplot(aes(x = GDP_PC, y = lifeExpectancy)) +
  geom_point(aes(size = totalPopulation, color = region))
gdple2019
```

### 3.

When comparing the graphs it can be seen that throughout the years GDP per capita and life expectancy has increased a substantial amount through the years.

Bigger countries have done better since the bigger data points have moved up which means that GDP PC and life expectancy has gone up. 

The Americas, Europe and Oceania have the highest GDP PC and life expectancy with African counties being in the bottom.

### 4.
```{r}
openfile %>% 
  filter(!is.na(region), !is.na(lifeExpectancy)) %>%
  filter(time == "1960" | time == "2019") %>% 
  group_by(region, time) %>%
  summarize(ave_lifeExpectancy = mean(lifeExpectancy))
```

### 5.
```{r}
openfile %>% 
  filter(!is.na(region), !is.na(lifeExpectancy)) %>%
  filter(time == "1960" | time == "2019") %>% 
  group_by(region, time) %>%
  summarize(avg_life = mean(lifeExpectancy)) %>% 
  mutate(growth_le = avg_life-lag(avg_life)) %>%
  arrange(-growth_le)
```

Asian and African have had the highest growth in life expectancy from 1960 to 2019, while European countries have seen the lowest change.

The Americas and Oceania have also experienced an increase in life expectancy.

### 6.
```{r}
openfile %>% filter(!is.na(GDP_PC), time == "1960" | time == "2019") %>%
  ggplot(aes(x = GDP_PC, fill = factor(time))) + geom_histogram(position = position_dodge())
```

### 7.
```{r}
openfile %>% filter(!is.na(lifeExpectancy), !is.na(name)) %>% filter(time == "1960" | time == "2019") %>%
  group_by(time) %>% mutate(rank = rank(desc(lifeExpectancy))) %>% filter(name == "United States of America") %>%
  reframe(name, time, rank)
```

### 8.
```{r}
openfile %>% 
  filter(!is.na(lifeExpectancy), !is.na(name)) %>%
  filter(time == "1960" | time == "2019") %>%
  group_by(time) %>% 
  mutate(rank = rank(desc(lifeExpectancy))) %>%
  mutate(countrybyyear = n_distinct(name)) %>% 
  mutate(ratio = rank/countrybyyear) %>% 
  reframe(name, time, ratio) %>% 
  filter(name == "United States of America")
```
This homework took me about 8-10 hours to complete